Formalization of Network Flow Algorithms

### 2017-02-14 paper/root.tex
paper final version
switched to condensed format
fixed affiliations
introduced the section about push-relabel algorithm

% Markierungen: TODO
\documentclass{llncs}

@@

\begin{document}

\title{Formalizing the Edmonds-Karp Algorithm}
% \subtitle{}

\author{Peter Lammich, S.~Reza Sefidgar}

\institute{Technische Universit\"at M\"unchen, \email{\{lammich,sefidgar\}@in.tum.de}}

\maketitle
\begin{abstract}
We present a formalization of the Ford-Fulkerson method for computing the maximum flow in a network.
Our formal proof closely follows a standard textbook proof, and is accessible even without being
an expert in Isabelle/HOL --- the interactive theorem prover used for the formalization.
We then use stepwise refinement to obtain the Edmonds-Karp algorithm, and formally prove a bound on its complexity.
Further refinement yields a verified implementation, whose execution time compares well to an unverified reference implementation in Java.

%We present a formalization of the Ford-Fulkerson method for computing the maximum flow in a network. 
%The formal proofs are done in Isabelle/HOL. They closely follow a standard textbook proof and are relatively easy 
%to read, even without being an Isabelle expert.

%We then use stepwise refinement to obtain the Edmonds-Karp algorithm and prove its upper complexity bound of $O(|V||E|^2)$.
%Further refinement yields a verified implementation, which is almost as fast as an unverified  
%reference implementation in Java.

% 
% and a verified imperative implementation of it. Our implementation is roughly 2.5 times slower than an unverified  
% reference implementation in Java. 
% 
% Finally, we prove the upper complexity bound of $O(|V||E|^2)$ for the Edmonds-Karp algorithm.
\end{abstract}

@@

For the capacity constraint, first observe that if $(u, v) \in E$, then $c_f(v, u) = f(u, v)$. Therefore, we have $f'(v,u) \leq c_f(v, u) = f(u, v)$, and hence
	\begin{align*}
	(f \uparrow f') (u, v) &= f(u, v) + f'(u, v) - f'(v, u)  && \text{(definition of $\augment$)} \\
	& \geq f(u, v) + f'(u, v) - f(v, u) && \text{(because $f'(v,u) \leq f(u, v)$)} \\
	& = f'(u, v) \\
	& \geq 0.
	\end{align*}
In addition,
	\begin{align*}
	(f \uparrow f') (u, v) &= f(u, v) + f'(u, v) - f'(v, u)  && \text{(definition of $\augment$)} \\
	& \leq f(u, v) + f'(u, v) && \text{(because flows are nonnegative)} \\
	& \leq f(u, v) + c_f(u, v) &&  \text{(capacity constraint)} \\
	& = f(u, v) + c(u, v) - f(u, v) && \text{(definition of $c_f$)} \\
	& = c (u, v).
	\end{align*}
% 
In the following we present the corresponding formal proof in Isar:
\Snippet{augment_flow_presv_cap}
The structure of the Isar proof is exactly the same as that of the textbook proof, except that we had to also consider the case $(u,v)\notin E$, which is not mentioned in the informal proof at all, and easily discharged in our formal proof by the \isai{auto}-tactic after the \isai{qed}. We also use exactly the same justifications as the original proof, except that we had to use the fact that there are no parallel edges to show $c_f(v,u)=f(u,v)$, which is not mentioned in the original proof.

@@

@@

%   
%   can be performed in two directions. First of all, the abstract theory can be extended to also support more advanced 
%   maximum flow algorithms like push-relabel algorithms~\cite{GoTa88}
%   
%   
%   
%   What do the Coq-Guys have: CFML? Others?
%   
%   
%   XXX, ctd here
%   
%   Unfortunately, it
% 
%   ... and related work
%     Mizar-Formalization: What did they do.
%     
%     History of the formalization: 
%       Impl: From first impl that was hardly able to compute a 100 nodes graph, to current efficient one.
%     
%     Future Work: Dinic. (Actually, our abstract scheme [almost] covers Dinic's algorithm!)
%     
% 
% 
%   
% Contributions
% 
%   Formal proof of mincut maxflow
%   fofo-scheme
%   inst to edmonds karp
%   complexity analysis of edmonds karp
%   refinement down to executable code. Roughly 2.5 times slower than Java. (What about OCaml)
%     + NetCheck
%     
% What shines (its a Pearl)
%   Min-Cut Max-Flow: Textboook like formal reasoning: Comprehensible proof, BUT machine checked
%     (Present one (carefully worked) example in paperI. We could use lemma \isai{augment_flow_presv_cap})
%     
%   Refinement based approach: Fofu-Scheme, instantiation to EdsKa. 
%     +1: Abstract Algo looks almost like pseudo-code you would expect in textbook.
%     +2: Fofu-Scheme proved correct for all aug-path finders. EdsKa is instantiation of it.
%     +3: Modularity: Fofu-scheme and pathfinder developed+proved independently of each other.
%   
%   Down to executable code, plugging in efficient data structures.
%   
% Some minor contributions:
%   Reusable BFS algorithm
%   Imperative matrix data structure (really minor).
%   
%   
%   
%     
% 
% 
%

>>>>

%%%%%%%%%%%%%%%%%%%%%%% file template.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
%\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%

@@

%
\begin{document}

\title{Formalization of Network Flow Algorithms}
\subtitle{a refinement approach in Isabelle/HOL}

%\titlerunning{Short form of title}        % if too long for running head

\author{Peter Lammich \and S. Reza Sefidgar}

%\authorrunning{Short form of author list} % if too long for running head

\institute{
  Peter Lammich \at
  Institut f\"ur Informatik,
  Technische Universit\"at M\"unchen \\
  \email{lammich@in.tum.de} %
\and
  S. Reza Sefidgar \at
  Institute of Information Security,
  Department of Computer Science,
  ETH Zurich, Switzerland \\
  \email{reza.sefidgar@inf.ethz.ch} %
}

\date{Received: date / Accepted: date}
% The correct dates will be entered by the editor

\maketitle

\begin{abstract}
We present a formalization of the Ford-Fulkerson method for computing the maximum flow in a network.
Our formal proof closely follows a standard textbook proof, and is accessible even without being
an expert in Isabelle/HOL --- the interactive theorem prover used for the formalization.
We then use stepwise refinement to obtain the Edmonds-Karp algorithm, and formally prove a bound on its complexity.
Further refinement yields a verified implementation, whose execution time compares well to an unverified reference implementation in Java.
\keywords{First keyword \and Second keyword \and More}
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

@@

For the capacity constraint, first observe that if $(u, v) \in E$, then $c_f(v, u) = f(u, v)$. Therefore, we have $f'(v,u) \leq c_f(v, u) = f(u, v)$, and hence
	\begin{align*}
	(f \uparrow f') (u, v) &= f(u, v) + f'(u, v) - f'(v, u)  && \text{(definition of $\augment$)} \\
	& \geq f(u, v) + f'(u, v) - f(u, v) && \text{(because $f'(v,u) \leq f(u, v)$)} \\
	& = f'(u, v) \\
	& \geq 0.
	\end{align*}
In addition,
	\begin{align*}
	(f \uparrow f') (u, v) &= f(u, v) + f'(u, v) - f'(v, u)  && \text{(definition of $\augment$)} \\
	& \leq f(u, v) + f'(u, v) && \text{(because flows are nonnegative)} \\
	& \leq f(u, v) + c_f(u, v) &&  \text{(capacity constraint)} \\
	& = f(u, v) + c(u, v) - f(u, v) && \text{(definition of $c_f$)} \\
	& = c (u, v).
	\end{align*}
% 
In the following we present the corresponding formal proof in Isar:
\Snippet{augment_flow_presv_cap}
The structure of the Isar proof is exactly the same as that of the textbook proof, except that we had to also consider the case $(u,v)\notin E$, which is not mentioned in the informal proof at all, and easily discharged in our formal proof by the \isai{auto}-tactic after the \isai{qed}. We also use exactly the same justifications as the original proof, except that we had to use the fact that there are no parallel edges to show $c_f(v,u)=f(u,v)$, which is not mentioned in the original proof.

@@

  \section{Push-Relabel algorithms}\label{sec:push_relabel}
  In this section we present the formalization of Push-relabel algorithms by
  Goldberg(XXX) that build up on Karzanov's idea of a preflow ---a relaxation
  of the flow definition by allowing the sum of flows over incoming edges to be
  greater equal than the sum of flows over outgoing edges of each vertex of the
  network. Karzanov was the first to leave the augmenting path idea in research
  on finding the maximum-flow in networks. He came up with an $O(V^3)$
  maximum-flow algorithm by relaxing some of the invariants in Dinitz's
  algorithm (XXX Dinitz paper). Years later, Goldberg proposed the generic
  Push-relabel algorithm which shares some ideas with Karzanov's algorithm, yet
  it provides freedom for choosing the order of ``push'' and ``relabel''
  operations. The push–relabel algorithm is considered one of the most efficient
  maximum flow algorithms. The generic algorithm has a strongly polynomial
  $O(V^2 E)$ time complexity, while specific variants of the algorithms achieve
  even lower time complexities up to $O(V^2 \sqrt{E})$. In what follows we first
  overview the transition from Ford-Fulkerson method to Push-relabel algorithm,
  and then provide the formal explanation of Goldberg's generic algorithm.

  One year after publication of Edmonds-Karp algorithm and without being aware
  of their result, Dinitz proposed another strongly polynomial algorithm based
  on Ford-Fulkerson method with $O(V^2 E)$ complexity. Original intention of
  Dinitz was just to accelerate iterations of Ford-Fulkerson method by means of
  a smart data structure. As finding the augmenting path is the bottleneck in each
  iteration of the Ford-Fulkerson method, he decided to keep track of all shortest
  paths from source to sink in a data-structure called the layered-network.
  Therefore, unless the length of the shortest path from source to sink has
  increased, we could use the layered-network for finding augmenting paths.
  Finding an augmenting path in the layered-network would require O(V)
  operations which is considerably more efficient than O(E) complexity of
  finding such path by searching the residual graph. Note that initializing the
  layered-network would has O(E) complexity, however this is only required
  whenever the length of the shortest path from source to sink has increased.

  Karzanov observed that at each phase of the Dinitz's algorithm it is
  sufficient to use a blocking-flow for increasing the current flow, and
  hence determining the next layered-network. He uses the term to refere to an
  augmenting flow in the current layared-network that saturates at least one edge
  on each path from source to the sink. During each phase, Karzanov's algorithm
  maintains a preflow in a layered-network. The algorithm ``pushes'' flow
  through the network to find a blocking-flow, and then ``balances'' the
  resulting preflow to obtain a valid flow at the end of each phase. Cherkarsky
  tired to eliminate balancing operation by canceling the previous goal of
  Karzanov's algorithm for constructing a feasible flow at the end of each
  phase. He proved that balancing the flow at some vertex is equivalent to
  pushing from it in future when it will appear in some layared-network at a
  greater distance from the sink. However, Cherkarsky could not come up with the
  generic push-relabel algorithm. The leftovers of phase structures of Dinitz's
  and Karsanov's algorithms prevented Cherkarsky from canceling the devision of
  his algorithm into phases.

  Goldberg's generic Push-relabel algorithm takes Karzanov's idea of using preflows
  to the next level. This algorithm abandons the idea of finding a flow in each
  phase, and also abandons the idea of global phases. Instead, Goldberg idea
  maintains a preflow in the original network and pushes flow toward the sink
  along what it estimates to be shortest paths. When the algorithm terminates,
  the preflow becomes a valid flow which is also proved to be a maximum flow.

  The Push-relabels algorithms maintain a preflow as well as a height-labeling
  during the execution. A preflow in a network with source $s$ and sink $t$ is a
  function $f: V \times V \mapsto \mathbb{R}$ that satisfies the capacity
  constraint as well as a relaxed conservation constraint\footnote{every flow
    related definition, like flow's value or flow's corresponding residual
    graph can be generalized for preflows.}. That is, for all vertices except
  the source $s$ the sum of flows over all incoming edges is greater equal than
  the sum of flows over all outgoing edges. The excess flow $e(v)$ of a vertex
  $v$ in the network is the amount by which in flow exceeds the out flow, and a
  $v$ is overflowing if $e(v) > 0$. Moreover, a height-labeling in a $s$-$t$
  network is a funtion $l: V \mapsto \mathbb{N}$ where $l(s) =
  \left\vert{V}\right\vert$, $l(t) = 0$, and for every edge $(u, v)$ of the
  residual graph we have that $l(u) \le l(v) + 1$.

  We summerize Cormen's intutition behind Push-relabel algorithms to clarify why
  such a labeling function is required. Consider vertices of the network to be
  liquid containers and residual network's edges to be pipes that are connecting
  those containers. The amount of flow that can be passed through each pipe
  corresponds to its residual capacity. Moreover, each of the containers is
  lying on a platform which elevates it to higher altitudes as the algorithm
  makes progress. The height-labeling function keeps track of the heights of
  each container, and the flow can only go from higher altitude containers to
  the containers at lower heights.

  At the beginning of the algorithm the source is at height
  $\left\vert{V}\right\vert$, and all other containers are on the ground (height
  $0$). The algorithm sends as much as flow downhill from the source to the
  sink, hence all the edges outgoing from source are filled to the capacity.
  This flow is stored in the intermediate containers (resembeling the idea of a
  preflow), and is eventually ``pushed'' to the sink. During the execution, it
  may happen that a non-empty container(resembeling the idea of an overflowing
  vertex) is only attached to pipes with left-over capacities that are
  connecting it to a containers at higher altitude. In such a case, the
  container is elevated (``relabeled'') to a higher altitude so it can push its
  excess flow. Eventually, all the flows that could get to the sink has reached
  there, and the intermediate containers has sent their liquids back to the
  source (resembeling a valid flow which is later proved to be also a maximum
  flow).

  \newcommand{\push}{{\mathbin\uparrow_P}}%
  As it was presented in the above intuition, Push-relabel algorithms perform
  two basic operations. The push operation $\push$ increases or decreases the preflow
  value of any edge $(u, v)$ of the network by pushing flow along its
  corresponding residual edges $(u, v)$ or $(v, u)$ respectively
  \footnote{applying push operation to an edge$(u, v)$ decreases $e(u)$ while
    increases $e(v)$ at the same time.}. However, it can be applied only if $u$
  is an overflowing vertex and $l(u) = l(v) + 1$. This restriction is due to the
  fact that no residual edges exists between two vertices with hight-labeling
  difference of more that 1. Hence, we would gain nothing by allowing such a
  push operation and the resulting preflow will remains unchanged. In the
  following definition, let $\Delta = min (e(u), c_f(u, v))$.
\[ (f \push (u,v)) (u', v') = 
  \begin{cases}
    f (u', v') \hfill & \text{ if $(u', v') \neq (u, v) \wedge (u', v') \neq (v, u)$}\\
    f (u', v') + \Delta & \text{ if $(u, v) \in E$}\\  
    f (v', u') - \Delta & \text{ if $(v, u) \in E$}\\
    0 & \text{ otherwise}
  \end{cases}
\]

Goldberg distinguishes two types of pushes which are latered used for proving
the correctness and complexity of his algorithm. A push is called saturating if
the edge $(u, v)$ is removed from the residual graph after the push operation,
i.e. $e(u) \ge c_f(u, v)$. Otherwise, the push is called non-saturating, and the
edge $(u, v)$ remains in the residual graph after the push operation.

  \newcommand{\relabel}{{\mathbin\uparrow_L}}%
Next, the relabel operation $\relabel$ which increases the height-labeling of an
overflowing vertex $u$, when for all edges $(u, v)$ that are going out of $u$ in
the residual graph we have $l(u) \le l(v)$. The overflowing precondition is
necessary to make sure that in the following definition minimization is not
applied to an empty set. Note that, according to the definition of the push
operation $\push$ flow can only be pushed downhill. So, with relabel operation
we enable an overflowing vertex to push its excess flow into one of its
neighbours. By definition, neither the source $s$ nor the sink $t$ can change
their height-labeling, so, the relabel operation does not apply to them.
\[ (l \relabel u) (u') = 
  \begin{cases}
    l(u') \hfill & \text{ if $u' \neq u$}\\
    1 + min \{l(v) \vert \text{$(u, v)$ is an edge of the residual graph}\}  & \text{ otherwise}
  \end{cases}
\]

The generic Push-relabel algorithm is then defined by iterative application of the aformentioned basic
operations. The algorithm starts by setting up an initial height-lableling and
preflow function. \isai{init-labeling} operation outputs a function $l$ such that
$l(s) = \vert V \vert$ and $l(u) = 0$ for any $u \neq s$. Also,
\isai{init-preflow} procedure produces a preflow function $f$ such that $f(s, u)
= c(s, u)$ for all edges going out of the source and $f(u, v) = 0$ for every
other edge of the network. The algorithm, terminates when no more push or
relabel operation is applicable, and according to correctness theorem of the
algorithm the resulting preflow upon termination is indeed a maximum-flow of the network.

\texttt{---Present either the refinement code or pseudocode---}

\texttt{---Discuss correctness and termination proofs if required}

\texttt{---Corelate to the FIFO and gap heuristic implementations}

\texttt{---Discuss our time complexity proof, if required}

@@

%\section{Introduction}
%\label{intro}
%Your text comes here. Separate text sections with
%\section{Section title}
%\label{sec:1}
%\subsection{Subsection title}
%\label{sec:2}
%as required. Don't forget to give each section
%and subsection a unique label (see Sect.~\ref{sec:1}).
%\paragraph{Paragraph headings} Use paragraph headings as needed.
%\begin{equation}
%a^2+b^2=c^2
%\end{equation}
%
%% For one-column wide figures use
%\begin{figure}
%% Use the relevant command to insert your figure file.
%% For example, with the graphicx package use
%  %\includegraphics{example.eps}
%% figure caption is below the figure
%\caption{Please write your figure caption here}
%\label{fig:1}       % Give a unique label
%\end{figure}
%%
%% For two-column wide figures use
%\begin{figure*}
%% Use the relevant command to insert your figure file.
%% For example, with the graphicx package use
% % \includegraphics[width=0.75\textwidth]{example.eps}
%% figure caption is below the figure
%\caption{Please write your figure caption here}
%\label{fig:2}       % Give a unique label
%\end{figure*}
%%
%% For tables use
%\begin{table}
%% table caption is above the table
%\caption{Please write your table caption here}
%\label{tab:1}       % Give a unique label
%% For LaTeX tables use
%\begin{tabular}{lll}
%\hline\noalign{\smallskip}
%first & second & third  \\
%\noalign{\smallskip}\hline\noalign{\smallskip}
%number & number & number \\
%number & number & number \\
%\noalign{\smallskip}\hline
%\end{tabular}
%\end{table}
%
%
%%\begin{acknowledgements}
%%If you'd like to thank anyone, place your comments here
%%and remove the percent signs.
%%\end{acknowledgements}
%
%% BibTeX users please use one of
%%\bibliographystyle{spbasic}      % basic style, author-year citations
%%\bibliographystyle{spmpsci}      % mathematics and physical sciences
%%\bibliographystyle{spphys}       % APS-like style for physics
%%\bibliography{}   % name your BibTeX data base
%
%% Non-BibTeX users please use
%%\begin{thebibliography}{}
%%
%% and use \bibitem to create references. Consult the Instructions
%% for authors for reference list style.
%%
%%\bibitem{RefJ}
%% Format for Journal Reference
%%Author, Article title, Journal, Volume, page numbers (year)
%% Format for books
%%\bibitem{RefB}
%%Author, Book title, page numbers. Publisher, place (year)
%% etc
%% \end{thebibliography}
